Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-24 paper/main.tex
do some small adjustments and extensions

\paragraph{Structure} Write structure. In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices. We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}. Finally, we will test our developments on an example from electronic structure calculation.

@@

\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
    \begin{equation}
        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
\end{theorem}

@@

@@

\todo{Summarize analysis}

@@

\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, $n_{\mtx{\Omega}}$ \emph{even}, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t < 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon
        \label{equ:chebyshev-nystrom-error}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all $\varepsilon > 0$ and $\delta \in (0, 1/2)$.
\end{theorem}

\begin{proof}
    The previously derived results can be applied to
    \begin{align}
        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p. $\geq 1 - \delta$)} \notag \\
        &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
        &\leq 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\
    \end{align}
\end{proof}

@@

We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
\begin{equation}
    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
    \label{equ:numerical-rank}
\end{equation}
$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$. When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
\begin{equation}
    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
    \label{equ:gaussian-kernel-eigenvalues}
\end{equation}
where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$. Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
\begin{equation}
    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
    \label{equ:gaussian-kernel-numerical-rank}
\end{equation}
with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants. The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$. This is illustrated in \reffig{fig:numerical-rank}.
\begin{figure}[ht]
    \centering
    \input{figures/numerical-rank.tex}
    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
        approximately computed by counting the number of eigenvalues
        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
    \label{fig:numerical-rank}
\end{figure}

@@

Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V.
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
The potential $V$ interacting with the electrons is generated by Gaussian wells
\begin{equation}
    v(r) = v_0 e^{-\lambda r^2}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $V_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf. \reffig{fig:gaussian-well}). The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$. This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian -- its spectral density -- allows us to interpret the system's energy levels.

\begin{figure}[ht]
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-1.pgf}
        \caption{$n_c=1$}
        \label{fig:gaussian-well-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-2.pgf}
        \caption{$n_c=2$}
        \label{fig:gaussian-well-2}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-5.pgf}
        \caption{$n_c=5$}
        \label{fig:gaussian-well-5}
    \end{subfigure}
    \caption{Periodic Gaussian well potential $V$ for different sizes of the supercell.}
    \label{fig:gaussian-well}
\end{figure}

>>>>

\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices. We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}. Finally, we will test our developments on an example from electronic structure calculation.

@@

\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, e^{-\sfrac{1}{2}}/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
    \begin{equation}
        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
\end{theorem}

@@

This theorem shows that for $\sigma \ll 1$, if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$.

@@

\todo{Summarize analysis (on one hand, half the degree $m$, but easier to interpolate because $\sigma \to \sqrt{2} \sigma$)}

@@

\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
        \label{equ:chebyshev-nystrom-error}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all $0 < \varepsilon \leq 1$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
\end{theorem}

\begin{proof}
    The previously derived results can be applied to
    \begin{align}
        &\int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \notag \\
        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
        &\leq \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi^{(m)}_{\sigma}(t) \right|~\mathrm{d}t && \text{(\refthm{thm:nystrom-pp} w.p. $\geq 1 - \delta$)} \notag \\
        &\leq (1 + \varepsilon) \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t + \varepsilon \int_{-1}^{1} \left| \phi_{\sigma}(t) \right|~\mathrm{d}t && \text{(triangle inequality)} \notag \\
        &\leq (1 + \varepsilon) 2 \underline{E}_{\sigma, m} + \varepsilon. && \text{(\reflem{lem:non-negative-chebyshev-error} and normalization)} \notag \\
    \end{align}
    From \reflem{lem:non-negative-chebyshev-error} we see that choosing $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$ ensures $\underline{E}_{\sigma, m} \leq \varepsilon$, and thus, since $\mathcal{\varepsilon} \leq 1$,
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq (1+\varepsilon) 2 \varepsilon + \varepsilon \leq 5 \varepsilon.
    \end{equation}
\end{proof}

@@

We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
\begin{equation}
    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
    \label{equ:numerical-rank}
\end{equation}
$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$.

%\begin{theorem}{Nyström estimator for parameter-dependent matrices}%{chebyshev-nystrom-raw}
%    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric %matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is %contained in $[-1, 1]$. Then the Chebyshev-Nyström++ estimate %$\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in %\mathbb{N}$, $n_{\mtx{\Omega}} \geq \max(6, \max_{t \in [-1, 1]}(r_%{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}))) + \log(\delta^%{-1}))$, $n_{\mtx{\Psi}} = 0$, and smoothing parameter $\sigma > 0$ %satisfies with for all $\gamma \geq 1$ with probability at least $1 %- \delta$
%    \begin{equation}
%        \int_{a}^{b} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}%(t) \right| ~\mathrm{d}t
%        \leq 2 e^2 (1 + r_{\varepsilon}) (\varepsilon + (n - r_%{\varepsilon}) \underline{E}_{\sigma, m}) + \underline{E}_%{\sigma, m}.
%    \end{equation}
%\end{theorem}
%
%\begin{proof}
%    \todo{TODO}
%\end{proof}

When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
\begin{equation}
    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
    \label{equ:gaussian-kernel-eigenvalues}
\end{equation}
where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$. Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
\begin{equation}
    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
    \label{equ:gaussian-kernel-numerical-rank}
\end{equation}
with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants. The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$. This is illustrated in \reffig{fig:numerical-rank}.
\begin{figure}[ht]
    \centering
    \input{figures/numerical-rank.tex}
    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
        approximately computed by counting the number of eigenvalues
        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
    \label{fig:numerical-rank}
\end{figure}

@@

Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
\begin{equation}
    v(r) = v_0 e^{-\lambda r^2}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf. \reffig{fig:gaussian-well}). The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$. This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian -- its spectral density -- allows us to interpret the system's energy levels.

\begin{figure}[ht]
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-1.pgf}
        \caption{$n_c=1$}
        \label{fig:gaussian-well-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-2.pgf}
        \caption{$n_c=2$}
        \label{fig:gaussian-well-2}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-5.pgf}
        \caption{$n_c=5$}
        \label{fig:gaussian-well-5}
    \end{subfigure}
    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes of the supercell.}
    \label{fig:gaussian-well}
\end{figure}
